Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-04 source/paper.tex
tidy figures
tidy some text

Compatible finite element methods provide a route to stable,
consistent discretisations of the partial differential equations
governing geophysical fluids \citep{cotter2023compatible}. They
possess many of the desirable properties described by
\citep{staniforth2012horizontal} as essential for the accuracy of the
dynamical core of weather and climate models, such as conservation of
mass (and potentially other quantities such as energy and enstrophy),
correct wave propagation speeds and avoidance of spurious
computational modes, even on non-orthogonal grids. Obtaining these
properties on a non-orthogonal grid is important because weather and
climate models achieve optimal performance on supercomputers by
performing calculations concurrently, or in parallel. In practice,
this means that the spatial domain is partitioned into subdomains that
are shared across the processors and data on the boundaries of the
subdomains needs to be shared between the processors to compute a
global solution. Current trends in supercomputer architectures

are run on massivly parallel supercomputers that require the co Spatial domain decomposition methods share the computation
across processors by While the required properties of the
discretisation guide the choice of finite element spaces, there are
many options for both the spatial and temporal discretisation and it
is not clear which will be best for any particular problem. Gusto
provides a flexible, extensible, easy-to-use toolkit of options for
rapid prototyping of different spatial and temporal discretisations
based on the compatible finite element framework applied to
geophysical fluid dynamics equations relevant to numerical weather and
climate prediction. In this paper we introduce Gusto to the community,
summarise its current capabilities and outline our vision for future
development.

@@

\begin{figure}\label{fig:2D de Rham}
  \centering{\includegraphics{de_rham_fig_2d.pdf}}
\end{figure}

\begin{figure}\label{fig:3D de Rham}
  \centering{\includegraphics{de_rham_fig_3d.pdf}}
\end{figure}

@@

\begin{figure}
  \centering{\includegraphics[width=0.8\textwidth]{gusto_schematic.png}}
\end{figure}

>>>>

Compatible finite element methods provide a route to stable,
consistent discretisations of the partial differential equations
governing geophysical fluids \citep{cotter2023compatible}. They
possess many of the desirable properties described by
\citet{staniforth2012horizontal} as essential for the accuracy of the
dynamical core of weather and climate models, such as conservation of
mass (and potentially other quantities such as energy and enstrophy),
correct wave propagation speeds and avoidance of spurious
computational modes, even on non-orthogonal grids. Obtaining these
properties on a non-orthogonal grid is important because weather and
climate models achieve optimal performance on supercomputers by
performing calculations concurrently, or in parallel, and orthogonal
grids of a spherical manifold will necessarily have singular points
that lead to a degredation in parallel scaling. An example of this is
the commonly-used latitude-longitude grid where the singular points
are the poles.

% In practice, this means that the spatial domain is partitioned into
% subdomains that are shared across the processors and data on the
% boundaries of the subdomains needs to be shared between the
% processors to compute a global solution. Current trends in
% supercomputer architectures
% are run on massivly parallel supercomputers that require the co
% Spatial domain decomposition methods share the computation
% across processors by

While the required properties of the discretisation guide the choice
of finite element spaces, there are many options for both the spatial
and temporal discretisation and it is not clear which will be best for
any particular problem. Gusto provides a flexible, extensible,
easy-to-use toolkit of options for rapid prototyping of different
spatial and temporal discretisations based on the compatible finite
element framework applied to geophysical fluid dynamics equations
relevant to numerical weather and climate prediction. In this paper we
introduce Gusto to the community, summarise its current capabilities
and outline our vision for future development.

@@

\begin{figure}\label{fig:2D de Rham}
  \centering{\includegraphics{figures/de_rham_fig_2d.pdf}}
\end{figure}

\begin{figure}\label{fig:3D de Rham}
  \centering{\includegraphics{figures/de_rham_fig_3d.pdf}}
\end{figure}

@@

\begin{figure}
  \centering{\includegraphics[width=0.8\textwidth]{figures/gusto_schematic.png}}
\end{figure}
